---
title: "Teacher salary bonuses"
date: '`r format(Sys.time(), "%B %e, %Y")`'
---

Teachers in North Carolina Public schools earn a bonus of \$750 if the students in their school meet a standard called "expected growth." A summary statistic called "average growth" is computed for each school; the expected growth standard is met when this summary measure exceeds zero. (This is based on [actual econometric research](https://my.vanderbilt.edu/performanceincentives/files/2012/10/200803_Vigdor_TeacherBonusesNC.pdf).)

For this example, we'll use regression discontinuity techniques to see if failing to receive a bonus motivates teachers to improve their performance in the subsequent school year.

- Running variable = `avgrow05`
- Treatment = `expgrow`
- Outcome = `avgrow06`

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(haven)
library(broom)
library(rdrobust)
library(rddensity)
library(modelsummary)

schools <- read_stata("data/nc_schools.dta")
```

# Step 1: Rule based?

Yep.

# Step 2: Fuzzy or sharp?

```{r}
ggplot(schools, aes(x = avgrow05, y = expgrow, color = avgrow05 > 0)) +
  geom_point(size = 0.5, alpha = 0.5,
             position = position_jitter(width = 0, height = 0.25, seed = 1234)) +
  geom_vline(xintercept = 0)
```

```{r}
schools %>%
  group_by(expgrow, avgrow05 > 0) %>%
  summarize(count = n())
```

```{r}
schools_clean <- schools %>% 
  filter(avgrow05 <= 0 & expgrow == 0 |
           avgrow05 > 0 & expgrow == 1)
```


# Step 3: Check for discontinuity in running variable around cutpoint

```{r}
ggplot(schools_clean, aes(x = avgrow05, fill = factor(expgrow))) +
  geom_histogram(binwidth = 0.05, color = "white", boundary = 0) +
  geom_vline(xintercept = 0) +
  labs(x = "Growth in 2005", y = "Count", fill = "Received bonus")

```

```{r}
test_density <- rddensity(schools_clean$avgrow05, c = 0)
summary(test_density)
```

```{r}
plot_density_test <- rdplotdensity(rdd = test_density,
                                   X = schools_clean$avgrow05,
                                   type = "both")  # This adds both points and lines
```

# Step 4: Check for discontinuity in outcome across running variable

```{r}
ggplot(schools_clean, aes(x = avgrow05, y = avgrow06, color = factor(expgrow))) +
  geom_point(size = 0.5, alpha = 0.5) +
  # Add a line based on a linear model for the people scoring 70 or less
  geom_smooth(data = filter(schools_clean, avgrow05 <= 0), method = "lm") +
  # Add a line based on a linear model for the people scoring more than 70
  geom_smooth(data = filter(schools_clean, avgrow05 > 0), method = "lm") +
  geom_vline(xintercept = 0) +
  labs(x = "2005 growth", y = "2006 growth", color = "Received bonus") +
  coord_cartesian(xlim = c(-0.3, 0.3), ylim = c(-0.3, 0.3))
```

# Step 5: Measure the size of the effect

## Parametric estimation

```{r}
summary(rdbwselect(y = schools_clean$avgrow06, x = schools_clean$avgrow05, c = 0))
```

```{r}
schools_bw <- schools_clean %>% 
  filter(avgrow05 > -0.061 & avgrow05 < 0.061)

ggplot(schools_bw, aes(x = avgrow05, y = avgrow06, color = factor(expgrow))) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(data = filter(schools_bw, avgrow05 <= 0), method = "lm") +
  geom_smooth(data = filter(schools_bw, avgrow05 > 0), method = "lm") +
  geom_vline(xintercept = 0) +
  labs(x = "2005 growth", y = "2006 growth", color = "Received bonus") +
  coord_cartesian(ylim = c(-0.25, 0.25))

ggplot(schools_bw, aes(x = avgrow05, y = avgrow06, color = factor(expgrow))) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(data = filter(schools_bw, avgrow05 <= 0), method = "loess") +
  geom_smooth(data = filter(schools_bw, avgrow05 > 0), method = "loess") +
  geom_vline(xintercept = 0) +
  labs(x = "2005 growth", y = "2006 growth", color = "Received bonus") +
  coord_cartesian(ylim = c(-0.25, 0.25))
```

```{r}
model_simple <- lm(avgrow06 ~ avgrow05 + expgrow,
                   data = schools_bw)
tidy(model_simple, conf.int = TRUE)
```

## Nonparametric estimation

```{r}
model_rdrobust <- rdrobust(y = schools_clean$avgrow06, x = schools_clean$avgrow05, c = 0) 

summary(model_rdrobust)
```

```{r}
rdplot(y = schools_clean$avgrow06, x = schools_clean$avgrow05, c = 0)
```


# Step 6: Compare all the effects

The **broom** package doesn't support `rdrobust` objects (yet), but it's possible to add `tidy()` and `glance()` functions that work with them, which means you can (1) use things like `tidy()` and (2) use things like `modelsummary()`. [Get the code here](https://github.com/tidymodels/broom/issues/777#issuecomment-643251123).

```{r}
#' Tidy a `rdrobust` regression discontinuity design object
#'
#' @param x A `rdrobust` object
#' @param ... extra arguments (not used)
#' @examples
#' \dontrun{
#' x <- runif(1000,-1,1)
#' y <- 3 * x + 2 * (x>=0) + rnorm(1000)
#' mod <- rdrobust(y,x)
#' 
#' tidy(mod)
#' 
#' glance(mod)
#' }
#' @export
tidy.rdrobust <- function(object, ...){
    ret <- tibble(term = row.names(object$coef), 
                  estimate = object$coef[, 1], 
                  std.error = object$se[, 1], 
                  statistic = object$z[, 1],
                  p.value = object$pv[, 1], 
                  conf.low = object$ci[,1],
                  conf.high = object$ci[, 2])
    ret
}

#' Glance a `rdrobust` regression discontinuity design object
#'
#' @inheritParams tidy.rdrobust
#' @export
glance.rdrobust <- function(object, ...){
    ret <- tibble(nobs.left = object$N[1],
                  nobs.right = object$N[2],
                  nobs.effective.left = object$N_h[1],
                  nobs.effective.right = object$N_h[2],
                  cutoff = object$c,
                  order.regression = object$q,
                  order.bias = object$q,
                  kernel = object$kernel,
                  bwselect = object$bwselect)
    ret
}

```

```{r}
tidy(model_rdrobust)
modelsummary(list(model_simple, model_rdrobust))
```

