---
title: "Evaluation and the causal revolution"
date: "2022-08-22"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sHcnk9KcoQVzDywxWRzVWHC"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "BIhz1dz5Vws", "Introduction", "1", TRUE,
  "f-sOSXyeeJE", "Data science and public service", "ds-ps", FALSE,
  "Z2foqSMWnGU", "Evidence, evaluation, and causation (1)", "evidence-evaluation-causation", FALSE,
  "W5ChuEhAzrs", "Evidence, evaluation, and causation (2)", "evidence-evaluation-causation1", FALSE,
  "D4LoEctZxHE", "Class details", "class-details", FALSE
)
```

## Readings

- The [syllabus](/syllabus.qmd), [content](/content/index.qmd), [examples](/example/index.qmd), and [assignments](/assignment/index.qmd) pages for this class
- {{< fa book >}} Chapter 1 in *Impact Evaluation in Practice* [@wb-eval]
- {{< fa book >}} Chapters 1 and 2 in *The Effect* [@Huntington-Klein:2021]
- {{< fa brands youtube >}} DJ Patil, ["What Makes a Radical and Revolutionary Technology?"](https://www.youtube.com/watch?v=UOoMzaWOQJA) 
    - <small>(DJ Patil is the former Chief Data Scientist of the United States under President Obama. He gave this forum address at Brigham Young University on February 13, 2018.)</small>
- {{< fa arrow-up-right-from-square >}} Stephen Goldsmith, ["Next Generation of Public Employees Must Understand Data and Policy"](http://www.govtech.com/opinion/Next-Generation-Public-Employees-Must-Understand-Data-Policy.html)
- {{< fa arrow-up-right-from-square >}} Hadley Wickham, ["Data Science: How is it Different To Statistics?"](https://imstat.org/2014/09/04/data-science-how-is-it-different-to-statistics%E2%80%89/)


## Slides

:::{.callout-warning}
**Important!!**: In the "Class details" video, I say that there are three books for the class: the World Bank evaluation book, *'Metrics Matter*, and *Causal Inference: The Mixtape*. [That's not the case this semester](/syllabus.qmd#books), since we're using the new *The Effect* book, which replaces both *'Metrics Matter* and the *Mixtape*. So ignore that part of the video. The slides are updated with the correct books, though: [see here](/slides/01-slides.html#85).
:::

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/01-slides")
slide_tabs(slide_details, "/slides/01-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

:::{.callout-warning}
**Important!!**: In the "Class details" video, I say that there are three books for the class: the World Bank evaluation book, *'Metrics Matter*, and *Causal Inference: The Mixtape*. [That's not the case this semester](/syllabus/#books), since we're using the new *The Effect* book, which replaces both *'Metrics Matter* and the *Mixtape*. So ignore that part of the video. The slides are updated with the correct books, though: [see here](/slides/01-slides.html#85).
:::

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```


## In-class stuff

Here are all the materials we'll use in class:

- [**Session 1 in-person slides**](/slides/01-class.html) ([PDF](/slides/01-class.pdf))

RStudio labs:

- [RStudio.cloud project](https://rstudio.cloud/spaces/269500/content/4377348)
- [Project `.zip` file](/projects/01-class.zip)
- [Lab slides 1: Markdown and universal writing](/slides/01-class_01_markdown-writing.html) ([PDF](/slides/01-class_01_markdown-writing.pdf))
- [Lab slides 2: Getting started with R and RStudio](/slides/01-class_02-getting-started.html) ([PDF](/slides/01-class_02-getting-started.pdf))
- [Lab slides 3: Data basics](/slides/01-class_03_data-basics.html) ([PDF](/slides/01-class_03_data-basics.pdf))
- [Lab slides 4: Visualize data with ggplot2](/slides/01-class_04_visualize-data.html) ([PDF](/slides/01-class_04_visualize-data.pdf))
- [Lab slides 5: Transform data with dplyr](/slides/01-class_05_transform-data.html) ([PDF](/slides/01-class_05_transform-data.pdf))


```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sHcnk9KcoQVzDywxWRzVWHC

https://youtu.be/BIhz1dz5Vws
PMAP 8521 • (1) Evaluation and the causal revolution: (0) Introduction

Introduction to the class session on evaluation and the causal revolution

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/01-content/


https://youtu.be/f-sOSXyeeJE
PMAP 8521 • (1) Evaluation and the causal revolution: (1) Data science and public service

Discussion of the role of data science in the public and nonprofit sectors

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/01-content/

  
https://youtu.be/Z2foqSMWnGU
PMAP 8521 • (1) Evaluation and the causal revolution: (2.1) Evidence, evaluation, and causation

Discussion of the role of evidence-based policy and program evaluation

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/01-content/


https://youtu.be/W5ChuEhAzrs
PMAP 8521 • (1) Evaluation and the causal revolution: (2.2) Evidence, evaluation, and causation

Discussion of why causation is tricky (but also possible!) to measure

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/01-content/


https://youtu.be/D4LoEctZxHE
PMAP 8521 • (1) Evaluation and the causal revolution: (3) Class details

Overview of how the class works

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/01-content/
```
