---
title: "Theories of change and logic models"
date: "2022-09-05"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sHjKJRcD-ebQUbcaFhc0cQH"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "dwsCSqczNng", "Introduction", "1", TRUE,
  "XkHUQJp5fh0", "Reproducibility", "reproducibility", FALSE,
  "Uos8RumBrzE", "Program theories", "program-theories", FALSE,
  "8CJ7M2dTmF4", "Logic models & results chains", "logic-models", FALSE
)
```

## Readings

- {{< fa book >}} Chapter 2 in *Impact Evaluation in Practice* [@wb-eval]
- {{< fa book >}} Chapter 2 in *Evaluation: A Systematic Approach* [@RossiLipseyHenry:2019]. This is available on iCollege.
- {{< fa book >}} Chapter 3 in *Evaluation: A Systematic Approach* [@RossiLipseyHenry:2019]. This is available on iCollege.


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/03-slides")
slide_tabs(slide_details, "/slides/03-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```

```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sHjKJRcD-ebQUbcaFhc0cQH

https://youtu.be/dwsCSqczNng
PMAP 8521 • (3) Theories of change: (0) Introduction

Introduction to the class session on theories of change

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/03-content/


https://youtu.be/XkHUQJp5fh0
PMAP 8521 • (3) Theories of change: (1) Reproducibility

Discussion about the importance of reproducible data analysis practices

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/03-content/

  
https://youtu.be/Uos8RumBrzE
PMAP 8521 • (3) Theories of change: (2) Program theories

Discussion of program components, program impact theories, and the importance of basing programs on sound theories

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/03-content/


https://youtu.be/8CJ7M2dTmF4
PMAP 8521 • (3) Theories of change: (3) Logic models & results chains

Discussion of results chains, logic models, and the difference between logic models and impact theories 

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/03-content/
```
