---
title: "DAGs and potential outcomes"
date: "2022-09-19"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sHwqEeiG5xb3zAHcQCGWZUC"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "6n7Qicffb2k", "Introduction", "1", TRUE,
  "vARju899D3E", "do()ing observational causal inference", "dag-adjustment", FALSE,
  "0pOA8y3YASA", "Potential outcomes", "potential-outcomes", FALSE
)
```

## Readings

- {{< fa podcast >}} Prologue and at least one of the four acts from [This American Life, "Gardens of Branching Paths," episode #691, January 10, 2020](https://www.thisamericanlife.org/691/gardens-of-branching-paths)
- {{< fa book >}} Chapter 3 in *Impact Evaluation in Practice* [@wb-eval]
- {{< fa book >}} Chapters 8, 9, and 10 in *The Effect* [@Huntington-Klein:2021]

### Potential outcomes, ATEs, and CATEs example page

- The [example page on potential outcomes, ATEs, and CATEs](/example/po-ate-cate.qmd) shows how to use R to calculate ATEs and CATEs


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/05-slides")
slide_tabs(slide_details, "/slides/05-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```


## In-class stuff

Here are all the materials we'll use in class:

- [**Session 5 FAQ slides**](/slides/05-class.html) ([PDF](/slides/05-class.pdf))

### Confounding, unblocked

<div class="ratio ratio-16x9">
<video controls loop>
<source src="/slides/img/05-class/video-confounding-unblocked.mp4" type="video/mp4">
</video>
</div>

### Confounding, blocked

<div class="ratio ratio-16x9">
<video controls loop>
<source src="/slides/img/05-class/video-confounding-blocked.mp4" type="video/mp4">
</video>
</div>

### Mediation

<div class="ratio ratio-16x9">
<video controls loop>
<source src="/slides/img/05-class/video-mediation.mp4" type="video/mp4">
</video>
</div>


```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sHwqEeiG5xb3zAHcQCGWZUC

https://youtu.be/6n7Qicffb2k
PMAP 8521 • (5) DAGs and potential outcomes: (0) Introduction

Introduction to the class session on DAGs and potential outcomes

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/05-content/


https://youtu.be/vARju899D3E
PMAP 8521 • (5) DAGs and potential outcomes: (1) do()ing observational causal inference

Discussion of the do-operator, how experiments let you manipulate DAGs, and how do-calculus lets you transform do-based expressions into do-free versions

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/05-content/

  
https://youtu.be/0pOA8y3YASA
PMAP 8521 • (5) DAGs and potential outcomes: (2) Potential outcomes

Discussion and demonstration of how the potential outcomes framework works and how it connects to DAGs

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/05-content/
```
