---
title: "Threats to validity"
date: "2022-09-26"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sFJtz7egfUASfhhz6zuj7GZ"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "46v8l60nQnI", "Introduction", "1", TRUE,
  "N4XYXuw_IA0", "Construct validity", "construct-validity", FALSE,
  "IziQBDzUJ6E", "Statistical conclusion validity", "statistical-conclusion-validity", FALSE,
  "ssz2DTfrGhw", "Internal validity", "internal-validity", FALSE,
  "uK7YN6KTRT4", "External validity", "external-validity", FALSE
)
```


## Readings

- {{< fa file-pdf >}} Guido Imbens, "Statistical Significance, *p*-Values, and the Reporting of Uncertainty" [@Imbens:2021]. This is available on iCollege, and [it's also free here](https://doi.org/10.1257/jep.35.3.157).
- {{< fa arrow-up-right-from-square >}} Randall Munroe, ["Significant"](https://xkcd.com/882/)
- {{< fa arrow-up-right-from-square >}} Alexander Coppock, ["10 Things to Know About Statistical Power"](https://egap.org/resource/10-things-to-know-about-statistical-power/)
- {{< fa trophy >}} Play around with FiveThirtyEight, ["Hack Your Way To Scientific Glory"](https://projects.fivethirtyeight.com/p-hacking/)
- {{< fa book >}} Chapter 9 in *Impact Evaluation in Practice* [@wb-eval]
- {{< fa book >}} Chapter 11 in *The Effect* [@Huntington-Klein:2021]


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/06-slides")
slide_tabs(slide_details, "/slides/06-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```


## In-class stuff

Here are all the materials we'll use in class:

- [**Session 6 FAQ slides**](/slides/06-class.html) ([PDF](/slides/06-class.pdf))
- [Zilch!](/resource/zilch.qmd)
- [{{< fa brands r-project >}} Live R script](https://www.dropbox.com/s/akez7ssaiyt6um3/bayes-fun.R)
- [{{< fa file-csv >}} `restaurant_inspections.csv`](https://vincentarelbundock.github.io/Rdatasets/csv/causaldata/restaurant_inspections.csv)

```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sFJtz7egfUASfhhz6zuj7GZ

https://youtu.be/46v8l60nQnI
PMAP 8521 • (6) Threats to validity: (0) Introduction

Introduction to the class session on threats to validity

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/06-content/


https://youtu.be/N4XYXuw_IA0
PMAP 8521 • (6) Threats to validity: (1) Construct validity

Discussion of threats to construct validity

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/06-content/

  
https://youtu.be/IziQBDzUJ6E
PMAP 8521 • (6) Threats to validity: (2) Statistical conclusion validity

Discussion of threats to statistical conclusion validity

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/06-content/
  
  
https://youtu.be/ssz2DTfrGhw
PMAP 8521 • (6) Threats to validity: (3) Internal validity

Discussion of threats to internal validity

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/06-content/
  
  
https://youtu.be/uK7YN6KTRT4
PMAP 8521 • (6) Threats to validity: (4) External validity

Discussion of threats to external validity

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/06-content/
```
