---
title: "Randomization and matching"
date: "2022-10-03"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sGJURMOwN9tf9MNDN4t0JMz"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "m4nKVJ-2K3I", "Introduction", "1", TRUE,
  "-zjSixOybQE", "The magic of randomization", "magic-randomization", FALSE,
  "W0NyALrjLA4", "How to analyze RCTs", "rct-how", FALSE,
  "vHb2TYAze74", 'The "gold" standard', "gold-standard", FALSE,
  "jECz-SS-sqc", "Adjustment with matching", "matching", FALSE
)
```

## Readings

- {{< fa file-pdf >}} Andrew Heiss, "Causal Inference," Chapter 10 in *R for Political Data Science: A Practical Guide* (2020) (Ignore the exercises!). [Get the PDF here.](/files/10-causal-inference.pdf)
- {{< fa book >}} Chapter 4 in *Impact Evaluation in Practice* [@wb-eval]
- {{< fa book >}} Chapters 11 and 13 in *The Effect* [@Huntington-Klein:2021]. **Focus especially on section 13.3 about standard errors.** And skim chapter 11; it's an overview of regression, which we reviewed in session 2, but it also applies the language of DAGs to regression, so look for that specifically as you read.
- {{< fa podcast >}} Planet Money, ["Moving To Opportunity?,"](https://www.npr.org/2019/08/30/756028025/episode-937-moving-to-opportunity) episode 937
- {{< fa arrow-up-right-from-square >}} Aaron Carroll, ["Workplace Wellness Programs Don’t Work Well. Why Some Studies Show Otherwise,"](https://www.nytimes.com/2018/08/06/upshot/employer-wellness-programs-randomized-trials.html) *The Upshot*, August 6, 2018

### RCTs, matching, and inverse probability weighting

- The [example page on RCTs](/example/rcts.qmd) shows how to use R to analyze and estimate causal effects from RCTs
- The [example page on matching and inverse probability weighting](/example/matching-ipw.qmd) shows how to use R to close backdoors, make adjustments, and find causal effects from observational data using matching and inverse probability weighting


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/07-slides")
slide_tabs(slide_details, "/slides/07-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```


## In-class stuff

Here are all the materials we'll use in class:

- [**Session 7 FAQ slides**](/slides/07-class.html) ([PDF](/slides/07-class.pdf))
- [In-class `.zip` file](/projects/rcts-ipw.zip)
- [RStudio.cloud project](https://rstudio.cloud/spaces/269500/content/4645983)

Other helpful resources:

- ["The Impact of Mask Distribution and Promotion on Mask Uptake and COVID-19 in Bangladesh"](https://www.poverty-action.org/study/impact-mask-distribution-and-promotion-mask-uptake-and-covid-19-bangladesh)
- Macartan Humphreys, ["I saw your RCT and I have some worries! FAQs"](https://macartan.github.io/i/notes/rct_faqs.html)
- Darren Dahly, ["Out of balance: A perspective on covariate adjustment in randomized experiments"](https://statsepi.substack.com/p/out-of-balance)
- [Bayesian stats and decison making](/resource/bayes.qmd)
- [Standard errors](/example/standard-errors.qmd)
- [Unobserved confounding and sensitivity analysis](/example/confounding-sensitivity.qmd)


```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sGJURMOwN9tf9MNDN4t0JMz

https://youtu.be/m4nKVJ-2K3I
PMAP 8521 • (7) Randomization and matching: (0) Introduction

Introduction to the class session on randomization and matching

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/07-content/


https://youtu.be/-zjSixOybQE
PMAP 8521 • (7) Randomization and matching: (1) The magic of randomization

Discussion of how randomization allows for strong claims of causal inference

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/07-content/

  
https://youtu.be/W0NyALrjLA4
PMAP 8521 • (7) Randomization and matching: (2) How to analyze RCTs

Discussion and demonstration of how to analyze the results of an RCT using R

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/07-content/
  
  
https://youtu.be/vHb2TYAze74
PMAP 8521 • (7) Randomization and matching: (3) The "gold" standard

Discussion of why RCTs aren't really a true "gold" standard for causal inference

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/07-content/
  
  
https://youtu.be/jECz-SS-sqc
PMAP 8521 • (7) Randomization and matching: (4) Adjustment with matching

Discussion of matching and inverse probability weighting + how these methods allow you to close DAG backdoors

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/07-content/


https://www.youtube.com/playlist?list=PLS6tnpTr39sGv7RPHEZy8CDSdCliQOW8K

https://youtu.be/OoYgOwxUQgI
PMAP 8521 • Example: Matching and IPW with R: 1: Drawing a DAG

Demonstration of how to connect nodes in a DAG with dagitty to find backdoor associations

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/matching-ipw/


https://youtu.be/_FNrfFe99R0
PMAP 8521 • Example: Matching and IPW with R: 2: Creating an RStudio project

Demonstration of how to create a new RStudio project and get ready to analyze data

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/matching-ipw/


https://youtu.be/7NW7GbO44BY
PMAP 8521 • Example: Matching and IPW with R: 3: Naive (and wrong!) estimate

Demonstration of how to calculate an incorrect correlation-is-not-causation estimate of a causal effect with R

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/matching-ipw/


https://youtu.be/uGwSRnET8Sg
PMAP 8521 • Example: Matching and IPW with R: 4: Matching

Demonstration of how to use the MatchIt package in R to close DAG backdoors, generate synthetic treatment and control groups, and estimate causal effects from observational data

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/matching-ipw/


https://youtu.be/CKm1rZlAwuA
PMAP 8521 • Example: Matching and IPW with R: 5: Inverse probability weighting

Demonstration of how to use inverse probability weighting with R to close DAG backdoors and estimate causal effects from observational data

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/matching-ipw/
```
