---
title: "Difference-in-differences I"
date: "2022-10-10"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sHydbEoTK9DkyKV92-uE3r-"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "4y84N23Mx-o", "Introduction", "1", TRUE,
  "ZHBZtKL7zAg", "Quasi-experiments", "quasi-experiments", FALSE,
  "B060u1zGOWo", "Interactions & regression", "interactions-regression", FALSE,
  "0v1aE70FhsQ", "Two wrongs make a right", "two-wrongs", FALSE,
  "tDWoNTk8CS4", "Diff-in-diff assumptions", "diff-diff-assumptions", FALSE
)
```

## Readings

- {{< fa book >}} Chapter 7 in *Impact Evaluation in Practice* [@wb-eval]
- {{< fa book >}} Chapter 18 in *The Effect* [@Huntington-Klein:2021]


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/08-slides")
slide_tabs(slide_details, "/slides/08-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```


## In-class stuff

Here are all the materials we'll use in class:

- [**Session 8 FAQ slides**](/slides/08-class.html) ([PDF](/slides/08-class.pdf))
- [Session 8 R code](/projects/week-8.zip) (on [RStudio.cloud](https://rstudio.cloud/spaces/269500/content/4698821))


```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sHydbEoTK9DkyKV92-uE3r-

https://youtu.be/4y84N23Mx-o
PMAP 8521 • (8/9) Difference-in-differences I & II: (0) Introduction

Introduction to the class session on difference-in-difference analysis

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/08-content/


https://youtu.be/ZHBZtKL7zAg
PMAP 8521 • (8/9) Difference-in-differences I & II: (1) Quasi-experiments

Discussion of what makes something a quasi-experiment + overview of three common types: difference-in-differences, regression discontinuity, and instrumental variables

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/08-content/

  
https://youtu.be/B060u1zGOWo
PMAP 8521 • (8/9) Difference-in-differences I & II: (2) Interactions & regression

Discussion of how interaction terms work in regression 

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/08-content/
  
  
https://youtu.be/0v1aE70FhsQ
PMAP 8521 • (8/9) Difference-in-differences I & II: (3) Two wrongs make a right

Overview of the intuition behind difference-in-differences analysis

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/08-content/
  
  
https://youtu.be/tDWoNTk8CS4
PMAP 8521 • (8/9) Difference-in-differences I & II: (4) Diff-in-diff assumptions

Discussion of the main assumptions for difference-in-differences

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/08-content/


https://www.youtube.com/playlist?list=PLS6tnpTr39sHw3FevrihLn2Ly8pSCUUag

https://youtu.be/u5iEtpITL3s
PMAP 8521 • Example: Diff-in-diff with R: 1: Getting started

Demonstration of how to create a new RStudio project and get ready to analyze data

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/diff-in-diff/


https://youtu.be/15BprphMT1g
PMAP 8521 • Example: Diff-in-diff with R: 2: Load and clean data

Demonstration of how to load and clean some data for diff-in-diff analysis

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/diff-in-diff/


https://youtu.be/SkPLMBkB06o
PMAP 8521 • Example: Diff-in-diff with R: 3: Exploratory data analysis

Demonstration of how to visually explore some key diff-in-diff variables

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/diff-in-diff/


https://youtu.be/56KPL2_nSxQ
PMAP 8521 • Example: Diff-in-diff with R: 4: Diff-in-diff manually

Demonstration of how to calculate average before/after treatment/control values and then calculate the difference in differences

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/diff-in-diff/


https://youtu.be/2JctZFGIYWw
PMAP 8521 • Example: Diff-in-diff with R: 5: Diff-in-diff with regression

Demonstration of how to use interaction terms in regression to calculate a diff-in-diff estimate

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/diff-in-diff/
```
