---
title: "Instrumental variables II + Regression discontinuity II"
date: "2022-11-07"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sGrbGfvdp0Gob6qAghIdiLq"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "qDAFXXriFyE", "Introduction", "1", TRUE,
  "_Ue7blCyEXk", "Treatment effects and compliance", "treatment-effects", FALSE,
  "M5cuBBuqZys", "Randomized promotion", "encouragement", FALSE,
  "9xGeCGRvVgE", "Fuzzy regression discontinuity", "fuzzy-rdd", FALSE
)
```

## Readings

- The [example page on complier average causal effects](/example/cace.qmd) shows how to use R to disentangle complier average causal effects
- The [example page on fuzzy regression discontinuity](/example/rdd-fuzzy.qmd) shows how to use R to use instrumental variables in fuzzy regression discontinuity, both parametrically and nonparametrically


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/12-slides")
slide_tabs(slide_details, "/slides/12-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```


## In-class stuff

Here are all the materials we'll use in class:

- [**Week 12 FAQ slides**](/slides/12-class.html) ([PDF](/slides/12-class.pdf))
- [{{< fa brands r-project >}} Live R script](https://www.dropbox.com/s/f1a04qzvmhelwcs/synthetic-data-f22.R)


```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sGrbGfvdp0Gob6qAghIdiLq

https://youtu.be/qDAFXXriFyE
PMAP 8521 • (12) Instrumental variables II + Regression discontinuity II: (0) Introduction

Introduction to the class session on instrumental variables and fuzzy regression discontinuity

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/12-content/


https://youtu.be/_Ue7blCyEXk
PMAP 8521 • (12) Instrumental variables II + Regression discontinuity II: (1) Treatment effects and compliance

Discussion of compliers, always takers, and never takers + how to find complier average causal effects (CACE) by finding the intent-to-treat effect (ITT) and dividing it by the proportion of compliers

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/12-content/

  
https://youtu.be/M5cuBBuqZys
PMAP 8521 • (12) Instrumental variables II + Regression discontinuity II: (2) Randomized promotion

Discussion of how to use randomized promotion as an instrument to measure the causal effect of a universal program with self-selection

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/12-content/
  
  
https://youtu.be/9xGeCGRvVgE
PMAP 8521 • (12) Instrumental variables II + Regression discontinuity II: (3) Fuzzy regression discontinuity

Discussion of how to use an above/below threshold variable as an instrument in fuzzy regression discontinuity designs

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/12-content/
```
