---
title: "Choosing and planning ethical evaluations"
date: "2022-11-14"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sEkOTbIIpoXYu8czSTl1yIh"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "rMUkZ4PSInY", "Introduction", "1", TRUE,
  "n2Sbyx0hz6I", "Types of evaluations", "types-evaluations", FALSE,
  "bXYPjwSl-Wg", "Model- and design-based inference", "model-design-inference", FALSE,
  "XdWbgeeCwZ8", "Ethics and open science", "ethics-open-science", FALSE
)
```

## Readings

- {{< fa arrow-up-right-from-square >}} ["Generating random numbers"](/example/random-numbers.qmd)
- {{< fa arrow-up-right-from-square >}} ["The ultimate guide to generating synthetic data for causal inference"](/example/synthetic-data.qmd)
- {{< fa file-pdf >}} Paul Hünermund and Beyers Louw, ["On the Nuisance of Control Variables in Regression Analysis"](https://arxiv.org/abs/2005.10314) [@HunermundLouw:2020] (**click on the PDF link in the right sidebar**)
- {{< fa file-pdf >}} ["Types of Evaluation,"](https://www.cdc.gov/std/Program/pupestd/Types%20of%20Evaluation.pdf) National Center for HIV/AIDS, Viral Hepatitis, STD, and TB Prevention, Centers for Disease Control (CDC)
- {{< fa book >}} Chapters 11–13 in *Impact Evaluation in Practice* [@wb-eval]
- {{< fa book >}} Chapter 15 in *The Effect* [@Huntington-Klein:2021] (skim this)


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/13-slides")
slide_tabs(slide_details, "/slides/13-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```


## In-class stuff

Here are all the materials we'll use in class:

- [**Week 13 FAQ slides**](/slides/13-class.html) ([PDF](/slides/13-class.pdf))
- [{{< fa brands r-project >}} Live R script](https://www.dropbox.com/s/f1a04qzvmhelwcs/synthetic-data-f22.R)


```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sEkOTbIIpoXYu8czSTl1yIh

https://youtu.be/rMUkZ4PSInY
PMAP 8521 • (13) Choosing and planning ethical evaluations: (0) Introduction

Introduction to the class session on types of evaluations, types of causal inference, and ethical evaluations

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/13-content/


https://youtu.be/n2Sbyx0hz6I
PMAP 8521 • (13) Choosing and planning ethical evaluations: (1) Types of evaluations

Brief overview of other types of evaluations, including needs assessment, monitoring, and cost-benefit analysis

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/13-content/

  
https://youtu.be/bXYPjwSl-Wg
PMAP 8521 • (13) Choosing and planning ethical evaluations: (2) Model- and design-based inference

Discussion of the difference between model- and design-based causal inference and how to know when to use different approaches

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/13-content/
  
  
https://youtu.be/XdWbgeeCwZ8
PMAP 8521 • (13) Choosing and planning ethical evaluations: (3) Ethics and open science

Discussion of how to use best practices of open science and ethical research to undertake good, ethical evaluations

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/13-content/
```
