---
title: "IPW: Job training"
output: html_document
date: '2022-10-03'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Do people who quit smoking gain weight?

> The National Supported Work Demonstration project [was] a transitional, subsidized work experience program for four target groups of people with longstanding employment problems: ex-offenders, former drug addicts, women who were long-term recipients of welfare benefits, and school dropouts, many with criminal records. The program provided up to 12-18 months of employment to about 10,000 individuals at 15 locations across the country for four years. In ten of these sites -- Atlanta, Chicago, Hartford, Jersey City, Newark, New York, Philadelphia, Oakland, San Francisco, and Wisconsin, 6,600 eligible applicants were randomly assigned either to experimental groups (offered a job in supported work) or to control groups, and an evaluation was conducted on the effects of the Supported Work Program. ([See here](https://www.icpsr.umich.edu/web/ICPSR/studies/7865) for more details.)

In 1986, Robert LaLonde [analyzed the results](https://www.jstor.org/stable/1806062) from this experiment and later compared it to observational data. We're going to use his data to see if participating in the job training program increased incomes.


## Load data

```{r libraries-data, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)
library(ggdag)

lalonde <- read_csv("data/lalonde.csv")
```

**Causal variables**:

- Treatment/exposure: `treat` (Individual was assigned to the job training program, 1 = yes, 0 = no)
- Outcome: `re78` (Individual's income in 1978, in US dollars)

**Possible confounders**:

- `age`: age in years
- `educ`: education in number of years of schooling
- `race`: the individual's race/ethnicity, (Black, Hispanic, or White)
- `married`: an indicator for marital status (1 = married, 0 = not married)
- `nodegree`: an indicator for whether the individual has a high school degree (1 = no degree, 0 = degree)
- `re74`: income in 1974, in US dollars
- `re75`: income in 1975, in US dollars


## Check difference in outcome across treatment status

Is there a visual difference?

```{r}
ggplot(lalonde, aes(x = re78, fill = factor(treat))) +
  geom_density(alpha = 0.5)
```

```{r}
lalonde %>% 
  group_by(treat) %>% 
  summarize(avg_income = mean(re78))
```

```{r}
model_naive <- lm(re78 ~ treat, data = lalonde)
tidy(model_naive, conf.int = TRUE)
```


## Draw a DAG

```{r fig.height=5, fig.width=7}
set.seed(1234)

treat_income_dag <- dagify(
  # Specify causes of job training program and 1978 income:
  treat ~ educ + race + married + nodegree + re74 + re75,
  re78 ~ educ + race + married + nodegree + re74 + re75,
  # Specify causal question:
  exposure = "treat", 
  outcome = "re78"
)

ggdag_status(treat_income_dag)
```

## Treatment model

```{r}
treatment_model <- glm(treat ~ age + educ + race + married +
                         nodegree + re74 + re75,
                       data = lalonde,
                       family = binomial(link = "logit"))
tidy(treatment_model)
```


## Propensity scores and weights

```{r}
lalonde_wts <- augment(treatment_model, lalonde, type.predict = "response") %>% 
  rename(propensity = .fitted) %>% 
  mutate(ipw = (treat / propensity) + ((1 - treat) / (1 - propensity)))
```

Investigate your weights. They should be clustered around 1 (with an average of 1), without any huge values.

```{r}
ggplot(lalonde_wts, aes(x = ipw)) +
  geom_density(fill = "darkred")
```

```{r}
lalonde_wts %>% summarize(avg_wt = mean(ipw))
mean(lalonde_wts$ipw)
```


## Outcome model

```{r}
outcome_model <- lm(re78 ~ treat, data = lalonde_wts, weights = ipw)
tidy(outcome_model, conf.int = TRUE)
```

```{r warning=FALSE, message=FALSE, results="hide"}
library(brms)
library(broom.mixed)
library(tidybayes)

outcome_bayes <- brm(
  bf(re78 | weights(ipw) ~ treat),
  data = lalonde_wts,
  cores = 4
)
```

```{r}
tidy(outcome_bayes)
```

```{r}
get_variables(outcome_bayes)

outcome_draws <- outcome_bayes %>% 
  gather_draws(b_treat)

ggplot(outcome_draws, aes(x = .value)) +
  stat_halfeye() +
  geom_vline(xintercept = 0)

outcome_draws %>% 
  summarize(prob_gt_0 = sum(.value > 0) / n())
```
