---
title: "IPW: Job training"
output: html_document
date: '2022-10-03'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Do people who quit smoking gain weight?

> The National Supported Work Demonstration project [was] a transitional, subsidized work experience program for four target groups of people with longstanding employment problems: ex-offenders, former drug addicts, women who were long-term recipients of welfare benefits, and school dropouts, many with criminal records. The program provided up to 12-18 months of employment to about 10,000 individuals at 15 locations across the country for four years. In ten of these sites -- Atlanta, Chicago, Hartford, Jersey City, Newark, New York, Philadelphia, Oakland, San Francisco, and Wisconsin, 6,600 eligible applicants were randomly assigned either to experimental groups (offered a job in supported work) or to control groups, and an evaluation was conducted on the effects of the Supported Work Program. ([See here](https://www.icpsr.umich.edu/web/ICPSR/studies/7865) for more details.)

In 1986, Robert LaLonde [analyzed the results](https://www.jstor.org/stable/1806062) from this experiment and later compared it to observational data. We're going to use his data to see if participating in the job training program increased incomes.


## Load data

```{r libraries-data, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)
library(ggdag)

lalonde <- read_csv("data/lalonde.csv")
```

**Causal variables**:

- Treatment/exposure: `treat` (Individual was assigned to the job training program, 1 = yes, 0 = no)
- Outcome: `re78` (Individual's income in 1978, in US dollars)

**Possible confounders**:

- `age`: age in years
- `educ`: education in number of years of schooling
- `race`: the individual's race/ethnicity, (Black, Hispanic, or White)
- `married`: an indicator for marital status (1 = married, 0 = not married)
- `nodegree`: an indicator for whether the individual has a high school degree (1 = no degree, 0 = degree)
- `re74`: income in 1974, in US dollars
- `re75`: income in 1975, in US dollars


## Check difference in outcome across treatment status

Is there a visual difference?

```{r}
# Density plot
```

```{r}
# group_by + summarize
```

```{r}
# Naive model
```


## Draw a DAG

```{r fig.height=5, fig.width=7}
set.seed(1234)

treat_income_dag <- dagify(
  # Specify causes of job training program and 1978 income:
  treat ~ educ + race + married + nodegree + re74 + re75,
  re78 ~ educ + race + married + nodegree + re74 + re75,
  # Specify causal question:
  exposure = "treat", 
  outcome = "re78"
)

ggdag_status(treat_income_dag)
```

## Treatment model

```{r}
# Treatment model
```


## Propensity scores and weights

```{r}
# Propensity scores and weights
```

Investigate your weights. They should be clustered around 1 (with an average of 1), without any huge values.

```{r}
# Check weights
```

```{r}
# Check average of weights
```


## Outcome model

```{r}
# Outcome model
```
