---
title: "Manipulating data with dplyr"
output: html_document
---


```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)  # For ggplot, dplyr, and friends

gapminder <- read_csv("data/gapminder.csv")
```

## gapminder

```{r}
gapminder
```


## Your turn 1

- `filter()` selects rows based on logical tests

The data for Canada:

```{r}
filter(gapminder, country == "New Zealand")
```

All data for countries in Oceania:

```{r}
filter(gapminder, country == "New Zealand")
```

Rows where the life expectancy is greater than 82:

```{r}
filter(gapminder, country == "New Zealand")
```


## Your turn 2

Canada before 1970:

```{r}
filter(gapminder, country == "New Zealand", year > 2000)
```

Countries where life expectancy in 2007 is below 50:

```{r}
filter(gapminder, country == "New Zealand", year > 2000)
```

Countries where life expectancy in 2007 is below 50, and are not in Africa:

```{r}
filter(gapminder, country == "New Zealand", year > 2000)
```


## Your turn 3

Write code to:

- Add an `africa` column that is TRUE if the country is on the African continent
- Add a column for logged GDP per capita (hint: use `log()`)
- Add an `africa_asia` column that says “Africa or Asia” if the country is in Africa or Asia, and “Not Africa or Asia” if it’s not

```{r}
mutate(gapminder)
```


## Your turn 4

Use `summarize()` to compute four statistics about the data:

- The first (minimum) year in the dataset
- The last (maximum) year in the dataset
- The number of rows in the dataset (use the cheatsheet)
- The number of distinct countries in the dataset (use the cheatsheet)

```{r}
gapminder 
```


## Your turn 5

Use `filter()` and `summarize()` to calculate the (1) the number of unique countries and (2) the median life expectancy on the African continent in 2007.

```{r}
gapminder 
```


## Your turn 6

Find the minimum, maximum, and median life expectancy for each continent.

```{r}
gapminder 
```

Find the minimum, maximum, and median life expectancy for each continent in 2007 only.

```{r}
gapminder
```
