---
title: "Effect of lowering the minimum legal drinking age on mortality"
output: html_document
---

## Load and clean data

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)
library(fixest)
library(tigris)

deaths_raw <- read_csv("data/deaths.csv")
```

The original data has a column called `fips` that contains id numbers for each state based on the Federal Information Processing Standards (FIPS) system ([see this](https://en.wikipedia.org/wiki/Federal_Information_Processing_Standard_state_code) for a list of all state FIPS codes). We can add state names and abbreviations to the dataset by merging a lookup table of FIPS codes (`fips_codes`) from the **tigris** package:

```{r}
fips <- fips_codes %>% 
  mutate(state_code = as.numeric(state_code)) %>% 
  group_by(state_code) %>% 
  slice(1) %>% 
  select(-county, -county_code)
```

```{r}
deaths <- deaths_raw %>% 
  mutate(state = as.numeric(state)) %>% 
  left_join(fips, by = c("state" = "state_code"))
```


## Simple diff-in-diff with Alabama and Arkansas

```{r}
# TODO: Make a new dataset based on `deaths` that only contains Alabama and Arkansas
# TODO: Make a new column called `after_change` that is TRUE (or 1) if year >= 1975
# TODO: Make a new column called `is_alabama` that is TRUE (or 1) if the state is Alabama
# TODO: Filter the data so that it (1) only includes years <= 1983, (2) only includes the 18-20 age group (agegr), and (3) includes only the death type (dtype) of "all"
```

```{r}
# TODO: Use group_by() and summarize() to find the average mortality rate (mrate) in Alabama and Arkansas before and after the change (after_change)
```

```{r}
# TODO: Use lm() to calculate the diff-in-diff effect with this model:
```

$$
\begin{aligned}
\text{Mortality}\ =&\ \beta_0 + \beta_1\ \text{Alabama} + \beta_2\ \text{After 1975}\ + \\
&\ \beta_3\ (\text{Alabama} \times \text{After 1975})
\end{aligned}
$$

```{r}
# TODO: Using the data for only Alabama and Arkansas, create a plot that shows
# the mortality rate over time, colored by state. Add a vertical line at 1975.
# Determine if the two states' trends are parallel pre-1975.
```


## TWFE with all states

```{r}
# TODO: Create a new dataset based on `deaths_raw` that (1) only includes years
# <= 1983, (2) only includes the 18-20 age group (agegr), and (3) includes only
# the death type (dtype) of "all"
```

```{r}
# TODO: Use that new data to create a TWFE model like this:
#
# HINT: The year column is currently numeric (like 1981, 1982, etc.), and R
# thinks that it's truly numeric (with possible values like 1980.432, etc.). To
# treat the year column like a categorical fixed effect, you need to make it a
# factor, so use factor(year) in your model code (or use mutate() above where
# you filter the data to create a factor version of the year column)
```

$$
\text{Mortality}\ =\ \beta_0 + \beta_1\ \text{Treatment} + \beta_2\ \text{State} + \beta_3\ \text{Year}
$$

```{r}
# TODO: Create another TWFE model that incorporates state-specific time trends by including the interaction of state and year:
```

$$
\begin{aligned}
\text{Mortality}\ =\ & \beta_0 + \beta_1\ \text{Treatment} + \beta_2\ \text{State} + \beta_3\ \text{Year} +\\
&\beta_4\ (\text{State} \times \text{Year})
\end{aligned}
$$
